---
title: "Data Visualization"
author: "David Svancer"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = "")
```


In this tutorial, we will learn about data visualization with `ggplot2`. 


We will be working with the `heart_disease` data set, where each row represents a patient from a cardiology department of a hospital and their associated medical test results.

The code below will load the `tidyverse` package and read the `heart_disease` data into our `R` session.


```{r}
library(tidyverse)

heart_df <- read_rds(url('https://gmudatamining.com/data/heart_disease.rds'))

```


# Data

Take a moment to explore the data below.


## Heart Disease Data

```{r}
heart_df
```


# Data Visualization

This section will cover the basics of data visualization with `ggplot2`. Most of the data visualizations that you will be required to produce follow the same general template, displayed below. 

To create different types of graphics with `ggplot2`, such as scatter plots or bar graphs, you will only need to adjust the input in brackets with the appropriate input or functions.

`ggplot2` makes use of the `+` operator. This is **not** the addition operator that is used in base `R`. 

Since `ggplot2` was created before `dplyr`, the `+` operator was used instead of the `%>%` operator, but within `ggplot2`, the `+` operator functions the same way as `%>%` functions in `dplyr`. 


```{r eval = FALSE}

ggplot(data = [DATA], mapping = aes([MAPPING]) +
         [GEOM_FUNCTION]() +
         [FACET_FUNCTION]() +
         [COORDINATE_FUNCTION]() +
         labs(title = [TITLE],
              x = [X AXIS LABEL],
              y = [Y AXIS LABEL])
       
```


We will be creating data visualizations using the `heart_df` data frame. Each row in this data frame represents a patient and their outcomes on medical tests. Each patient eventually did or did not develop heart disease. This is captured by the `heart_disease` variable.


## Scatter Plots

### Simple Scatter Plot

I will introduce each bracketed parameter in the template above by building a series of scatter plots that adds on each layer. We will be using the Heart Disease data set.

The code below produces a simple scatter plot. The data parameter in `ggplot()` takes the data frame that contains your data. The `aes()` option in the mapping parameter stands for aesthetic mapping. 

Below we are telling `ggplot()` to map the `age` column values in `heart_df` to the x-axis and the `cholesterol` column values to the y-axis. This produces (x,y) pairs that are mapped to coordinates on the graph.

Finally, we add the `geom_point()` function which tells `ggplot` that you would like points to be displayed on the coordinate mappings you have created.


```{r}

ggplot(data = heart_df, mapping = aes(x = age, y = cholesterol)) +
       geom_point()

```

All `geom` functions take optional aesthetic values. These include, `color`, `size`, and `alpha` (to control opacity). When these parameters are placed within a `geom` function and outside of the `aes` argument, they are applied to **all points**. You can specify the name of a color in `R` such as "orange" or by supplying a HEX code as I did below.

Below, I am styling the points to have size 2, a blue color using the HEX code #006EA1, and 45% opacity.


```{r}

ggplot(data = heart_df, mapping = aes(x = age, y = cholesterol)) +
       geom_point(color = "#006EA1", size = 2, alpha = 0.45)

```


### Adding a Third Variable With `aes()`

Now we want to visualize the scatter plot by groups, those with heart disease and those without. 

We can color the points by the value of the `heart_disease` variable as follows. Add `color = heart_disease` inside of the `aes` function within the mapping. This adds to the original mapping that we provided to `ggplot()`. 

Now we have mapped each point to three attributes (`age`, `cholesterol`, color based on `heart_disease` value).


```{r}

ggplot(data = heart_df, mapping = aes(x = age, y = cholesterol, color = heart_disease)) +
       geom_point()

```


### Adding a Third Variable by Facet

We can also visualize a third variable by adding the `facet_wrap()` function. 

This function separates the scatter plot based on the value of `heart_disease`.  

The general syntax is `facet_wrap( ~ Your_facet_variable, nrow = rows_you_would_like)`.


```{r}

ggplot(data = heart_df, mapping = aes(x = age, y = cholesterol)) +
       geom_point() +
       facet_wrap(~ heart_disease, nrow = 2)

```


If you would like to keep the `heart_disease` values as different colors, just add `color = heart_disease` into `aes`. 

Below I demonstrate what changing the `nrow` option will do to your plot.


```{r}

ggplot(data = heart_df, mapping = aes(x = age, y = cholesterol, color = heart_disease)) +
      geom_point() +
      facet_wrap(~ heart_disease, nrow = 1)

```


### Adding a Fourth Variable by Facet

We can visualize the scatter plot by combinations of character or factor variables. If you are faceting by more than one variable, use the `facet_grid()` function. 

The general syntax for `facet_grid()` is `facet_grid(Vertical_variable ~ Horizontal_variable)`


```{r}

ggplot(data = heart_df, mapping = aes(x = age, y = cholesterol, color = heart_disease)) +
       geom_point() +
       facet_grid(heart_disease ~ fasting_blood_sugar)

```

Finally, let's change the title and labels of our plot.


```{r}

ggplot(data = heart_df, mapping = aes(x = age, y = cholesterol, color = heart_disease)) +
  geom_point() +
  facet_grid(heart_disease ~ fasting_blood_sugar) +
  labs(title = "Cholesteral vs Age by Heart Disease and Fasting Blood Sugar Levels",
       x = "Patient Age",
       y = "Patient Cholesterol")

```


## Bar Charts

Now that we have gone through the layers of the template, let's see what adjustments are needed to create a bar chart. 

Let's visualize the number of patients with and without heart disease. Now our `aes` mapping in `ggplot()` only has one variable, heart_disease. 

To produce a bar chart, we use the `geom_bar()` function. The `stat = "count"` option tells `ggplot()` to transform the `heart_disease` column values and generate the counts for each level, Yes or No.


```{r}

ggplot(data = heart_df, mapping = aes(x = heart_disease)) +
       geom_bar(stat = "count")

```


In the graph above, behind the scenes `ggplot()` is computing the following transformation to obtain the counts for each level of `heart_disease`.



```{r}

heart_df %>% group_by(heart_disease) %>% 
             summarise(count = n())

```


If you wanted to compute the summary statistics yourself, you could do the following. 

Using `dplyr`, create a summary data frame, `heart_summary`, that has two variables, `heart_disease` and `count`. The `count` variable will have the number of patients that belong to the corresponding value of `heart_disease`.

Next, you must add `y = count` into the `aes` option in `ggplot()`. This tells `ggplot()` that you want to plot two pairs of coordinates (heart_disease value, count for heart_disease value). In this example, (No, 160), and (Yes, 137).

Finally, you must change `stat = "count"` in `geom_bar()` to `stat = "identity"`


```{r}

heart_summary <-  heart_df %>% 
                  group_by(heart_disease) %>% 
                  summarise(count = n())
```


```{r}
#View results
heart_summary

```


```{r}
# Plot the data, same as before
ggplot(data = heart_summary, mapping = aes(x = heart_disease, y = count)) +
       geom_bar(stat = "identity")

```


Let's use this same methodology to plot a bar chart of average `old_peak`, by `chest_pain`. This time, let's make the bars blue with a white border and add text labels. To change the bar color for all bars, use the `fill` option within `geom_bar()`. To change the border color, use `color`.


```{r}

oldpeak_summary <- heart_df %>% group_by(chest_pain) %>% 
                   summarise(avg_oldpeak = mean(old_peak))
```


```{r}
# View results
oldpeak_summary

```



```{r}
# Plot the data
ggplot(data = oldpeak_summary, mapping = aes(x = chest_pain, y = avg_oldpeak)) +
      geom_bar(stat = "identity", fill = "#006EA1", color = "white") +
      labs(title = "Average Old Peak by Chest Pain",
           x = "Chest Pain",
           y = "Average Old Peak")

```


### Reorder The Categories of a Bar Chart with *reorder()*

To reorder the categorical values of either the x or y axis on a bar chart, we can use the `reorder()` function from base `R`. 

The `reorder()` function has two required arguments - a vector of values in the first argument and another vector of values of equal length as the second argument by which to sort the first. Let's see some examples below.


```{r}

 ggplot(data = oldpeak_summary, mapping = aes(x = reorder(chest_pain, avg_oldpeak), 
                                              y = avg_oldpeak)) +
        geom_bar(stat = "identity", fill = "#006EA1", color = "white") +
        labs(title = "Average Old Peak by Chest Pain",
             x = "Chest Pain",
             y = "Average Old Peak")

```


```{r}
# To sort values in reverse order, simply put a minus (-) in 
# front of the second variable
ggplot(data = oldpeak_summary, mapping = aes(x = reorder(chest_pain, -avg_oldpeak), 
                                              y = avg_oldpeak)) +
        geom_bar(stat = "identity", fill = "#006EA1", color = "white") +
        labs(title = "Average Old Peak by Chest Pain",
             x = "Chest Pain",
             y = "Average Old Peak")

```


Let's go back to our original bar chart that showed the number of patients with and without heart disease. If we wanted to color the bars by `heart_disease` value, we can perform the same step as with the scatter plot. Now we add `fill = heart_disease` into the `aes` mapping. 

When you add the `fill` option into `aes` it updates the aesthetic mapping to include three dimensions for each bar (`heart_disease` value, count, fill color of `heart_disease` value). 


```{r}

ggplot(data = heart_df, mapping = aes(x = heart_disease, fill = heart_disease)) +
       geom_bar(stat = "count") + 
       labs(title = "Heart Disease Prevalence", x = "Heart Disease Status",
            y = "Number of Patients")

```


### Adding a Facet Variable

We can add a facet variable in the same way as we did with the scatter plot example, by adding `facet_wrap()`


```{r}

ggplot(data = heart_df, mapping = aes(x = heart_disease, fill = heart_disease)) +
       geom_bar(stat = "count") + 
       facet_wrap(~ chest_pain, nrow = 1) +
       labs(title = "Heart Disease Prevalence by Chest Pain", x = "Heart Disease Status",
            y = "Number of Patients")

```

## Stacked Bar Charts

If we add a variable to `aes(fill = )` that is different from the variable mapped to the x-axis, then we create a stacked bar chart. 

The variable which we add to the `fill` option in `aes`, will add a color based on the level of that variable and will display its width by the number of observations in the data. 

Below we create a stacked bar chart by adding the `chest_pain` variable.


```{r}

ggplot(data = heart_df, mapping = aes(x = heart_disease, fill = chest_pain)) +
       geom_bar(stat = "count") + 
       labs(title = "Heart Disease Prevalence by Chest Pain",
            x = "Heart Disease Status",
            y = "Number of Patients")

```


We can also create a 100 percent stacked column chart by adding `position = "fill"` into `geom_bar()`. This will display the proportion of observations that fall into the various values of `chest_pain` within each `heart_disease` value.


```{r}

ggplot(data = heart_df, mapping = aes(x = heart_disease, fill = chest_pain)) +
       geom_bar(stat = "count", position = "fill") + 
       labs(title = "Heart Disease Prevalence by Chest Pain",
            x = "Heart Disease Status",
            y = "Proportion of Patients")

```


### Stacking Bars Side by Side

A third option for working with stacked bar plots is to add `position = "dodge"` within `geom_bar()` to stack bars side by side.

These type of plots are usually used to compare multiple values or proportions that are changing over time. 

Let's apply this style to our visualization of `heart_disease` and `chest_pain`.


```{r}

ggplot(data = heart_df, mapping = aes(x = heart_disease, fill = chest_pain)) +
       geom_bar(stat = "count", position = "dodge") + 
       labs(title = "Heart Disease Prevalence by Chest Pain",
            x = "Heart Disease Status",
            y = "Number of Patients")

```

## Column Charts

To obtain a column chart, we can first build a bar chart and then use the `coord_flip()` function to flip the x and y axes. In the example below, we first build a bar chart and then take the necessary steps to build a column chart.


### Build a Bar Chart

```{r}

ggplot(data = heart_df, aes(x = fasting_blood_sugar, fill = heart_disease)) +
    geom_bar(stat = "count") +
    labs(title = "Fasting Blood Sugar Level by Heart Disease Status",
         x = "Fasting Blood Sugar", y = "Number of Patients")
         
```


### Flip The Axes With `coord_flip()`


```{r}

ggplot(data = heart_df, aes(x = fasting_blood_sugar, fill = heart_disease)) +
    geom_bar(stat = "count") +
    labs(title = "Fasting Blood Sugar Level by Heart Disease Status",
         x = "Fasting Blood Sugar", y = "Number of Patients") +
    coord_flip()
         
```


## Histograms

Histograms are used to visualize the distribution of continuous variables. They automatically create a series of `bins` that combine the values of a numeric variable into categories. Then the number of times the original values fall into these bins is counted and displayed as a vertical bar. These graphs are used to visually assess the properties of numeric variables, such as symmetry, skewness, variability, and central tendency.

To create a histogram, use the `geom_histogram()` function. In the example below I have added `fill = "#006EA1"` and `color = "white"` inside the `geom_histogram()` function. The `fill` option specifies the bar color, while the `color` option specifies the border color of the bars, just as in bar charts.


```{r}

ggplot(data = heart_df, mapping = aes(x = resting_blood_pressure)) +
       geom_histogram(fill = "#006EA1", color = "white") + 
       labs(title = "Distribution of Resting Blood Pressure",
            x = "Resting Blood Pressure",
            y = "Number of Patients")

```


The default number of bins for histograms is set to 30. To change this, adjust the `bins` option within the `geom_histogram()` function. In the example below, the histogram is created using 15 bins for the `RestBP` variable.



```{r}

ggplot(data = heart_df, mapping = aes(x = resting_blood_pressure)) +
      geom_histogram(fill = "#006EA1", color = "white", bins = 15) + 
      labs(title = "Distribution of Resting Blood Pressure",
           x = "Resting Blood Pressure",
           y = "Number of Patients")

```


### Adding Additional Variables to Histograms

Let's look at the distribution of `resting_blood_pressure` by the `heart_disease` variable. We can do this by faceting. I demonstrate how to do this with one and two facet variables in the code below


```{r}

ggplot(data = heart_df, mapping = aes(x = resting_blood_pressure, fill = heart_disease)) +
       geom_histogram(color = "white", bins = 15) + 
       facet_wrap( ~ heart_disease, nrow = 1) +
       labs(title = "Distribution of Resting Blood Pressure",
            x = "Resting Blood Pressure",
            y = "Number of Patients")

```


We can use `facet_grid()` to add another variable to the visualization, `thalassemia`.


```{r}

ggplot(data = heart_df, mapping = aes(x = resting_blood_pressure, fill = heart_disease)) +
       geom_histogram( color = "white", bins = 15) + 
       facet_grid(heart_disease ~ thalassemia) +
       labs(title = "Distribution of Resting Blood Pressure",
            x = "Resting Blood Pressure",
            y = "Number of Patients")

```

## Density Histograms

Density histograms are used to determine whether a particular theoretical probability density function is a good fit to model the dynamics of that variable in a population. Unlike regular histograms, which provide counts of a numeric variable for a series of bins, density histograms provide the **proportion** of observations falling within a series of bins.

To create a density histogram, use the same syntax as when creating a standard histogram, with the exception of adding `y = ..density..` into the `aes()` option.


```{r}

# Density histogram of resting_blood_pressure
ggplot(data = heart_df, mapping = aes(x = resting_blood_pressure, y = ..density..)) +
       geom_histogram(fill = "#006EA1", color = "white", bins = 15) + 
       labs(title = "Distribution of Resting Blood Pressure",
            x = "Resting Blood Pressure",
            y = "Proportion of Patients")

```



## Boxplots

Box plots are a powerful tool for exploring the central tendency and variability of numeric data by various levels of a factor or character. In the examples below, we will produce boxplots with `ggplot()`. 

The `geom` function used is `geom_boxplot()`. 

For boxplots, the `x` argument in `aes` should be a character or factor variable. The "Box" in boxplots represents the IQR (25th - 75th percentiles of data values).

In the next couple of examples, let's use the built-in data frame, `mpg` which has data on over 200 cars and their associated features such as city and highway miles per gallon. 



```{r}
mpg
```


The boxplot below visualizes the distribution of `hwy` values by car `class`


```{r}

# Box plots of "hwy" variable by "class"
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot(fill = "#006EA1") +
  labs(title = "Boxplot of hwy by class", x = "Class of Vehicle",
       y = "Miles per Gallon Highway (hwy)")

```


 To color the boxplots by the level of a character or factor variable, we must add this to the aesthetic mapping `aes()`.
 
 In the example below, let's use the `reorder()` function to sort the categories on the x-axis. 
 
 The `reorder()` function takes an optional third argument, `FUN`, which is usually an aggregation function. 
 
 In the example below, we are reordering the `class` categories on the x-axis by the median value of `hwy` within each `class` category.



```{r}

ggplot(data = mpg, mapping = aes(x = reorder(class, hwy, FUN = median), 
                                 y = hwy, fill = class)) +
  geom_boxplot() + 
  labs(title = "Boxplot of hwy by class",
       x = "Class", 
       y = "Miles per Gallon Highway (hwy)")

```


In cases where the factor of character variable has long labels, it is useful to use the `coord_flip()` function to switch the order of the axes in the plot. This is demonstrated below.



```{r}
# Boxplot fill colors by "class" variable values and flip axes
ggplot(data = mpg, mapping = aes(x = reorder(class, hwy, FUN = median),
                                 y = hwy, fill = class)) +
  geom_boxplot() + 
  labs(title = "Boxplot of hwy by class", 
       x = "Class", y = "Miles per Gallon Highway (hwy)") + 
  coord_flip()

```


## Violin Plots With Data Points

A violin plot is another way to visualize numeric variables by character or factor categories to study the numeric value distributions. 

In `ggplot2`, the `geom_violin()` function produces these plots. It is similar to a box plot except that instead of displaying a box for the inter-quartile range (IQR), it display what's known as a `kernel density estimate`. 

A kernel density estimate is simply an estimated probability density function that is plotted along the vertical axis of the plot, symmetrically on both sides. 

The more this function sticks out (greater width), the more the original data points are located in that region of the y-axis.

This visualization is best combined with `geom_jitter()`, which overlays the original data points onto the vertical axis. The `width` argument gives the amount by which to scatter the data points and the `alpha` argument provides the level of color saturation (0.5 represents 50% of the completely filled data points).


```{r}

ggplot(data = mpg, mapping = aes(x = reorder(class, hwy, FUN = median), 
                                 y = hwy, fill = class)) +
  geom_violin() +  
  geom_jitter(width = 0.07, alpha = 0.5) +
  labs(title = "Violin Plot of hwy by class",
                        x = "Class", y = "Miles per Gallon Highway (hwy)")

```


## Line Charts

To create line charts, we use the `geom_line()` function.

Let's say we are interested in studying the average maximum heart rate (`max_heart_rate` variable) and by patient `age`. First, I will use `dplyr` to create a summary data frame with this information.


```{r}

heart_summary <-  heart_df %>% group_by(age) %>% 
                  summarise(patients = n(),
                            avg_max_hr = mean(max_heart_rate)) %>% 
                  arrange(age) %>% 
                  filter(patients >= 5) # Keep Ages with at least 5 patients
```



```{r}
# Let's take a look at the results
heart_summary

```


### Simple Line Chart

Let's plot a line chart with `avg_max_hr` on the y-axis and `age` on the x-axis

```{r}

ggplot(data = heart_summary, mapping = aes(x = age, y = avg_max_hr)) +
       geom_line(color = "#0072B2")

```


Now let's include points in the line chart.


```{r}

ggplot(data = heart_summary, mapping = aes(x = age, y = avg_max_hr)) +
       geom_line(color = "#0072B2") +
       geom_point(color = "#0072B2")

```


Finally, if you are not a fan of the default grey background in `ggplot2`, just add `theme_light()` to the end of any plot. There are many more themes available in `ggplot2`, such as theme_bw(), or theme_classic(). Feel free to experiment with them. 


```{r}

ggplot(data = heart_summary, mapping = aes(x = age, y = avg_max_hr)) +
  geom_line(color = "#0072B2") +
  geom_point(color = "#0072B2") +
  theme_light()

```


## Learning More

To learn more about the advanced data visualization capabilities of `ggplot2` please see the following resources:

- [Data Visualization with R](https://rkabacoff.github.io/datavis/)
- [R for Data Science - Data Visualization](https://r4ds.had.co.nz/data-visualisation.html)
- [ggplot2 documentation](https://ggplot2.tidyverse.org/)



