---
title: "Strings, Dates, Factors, and Iteration"
author: "David Svancer"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = "")
```


In this tutorial, we will cover the following topics:

- Working with categorical data and factors using the `forcats` package
- Creating and manipulating dates with the `lubridate` package
- Manipulating text data with the `stringr` package
- Making decisions in programming with the `if_else()` function
- Iteration with the `purrr` package to apply functions to multiple elements of a list or data frame


First let's load the `tidyverse` and `lubridate` packages as well as the `employee_data` and `home_sales` tibbles into our `R` environment.


```{r}
library(tidyverse)
library(lubridate)
```

```{r}
employee_data <- read_rds(url('https://gmudatamining.com/data/employee_data.rds'))
```

```{r}
home_sales <- read_rds(url('https://gmudatamining.com/data/home_sales.rds'))
```

## Data

We will be working with the `employee_data` data set, where each row represents an employee who either did or did not resign from a company as well as their attributes and work history, and the `home_sales` data, where each row represents a real estate home sale in the Seattle area between 2014 and 2015.

Take a moment to explore these data sets below.

### Employee Data

```{r}
employee_data
```


### Seattle Home Sales

```{r}
home_sales
```

# Categorical Data

Categorical data, sometimes referred to as nominal data or ordinal data, represents data that only differs by label (such ‘yes’/‘no’) or ranks (such as ‘1st’, ‘2nd’, etc.).

For example, the `department` variable in our `employee_data` is a categorical variable with the following levels: Marketing, Sales, Research, Product Development, IT and Analytics, and Finance and Operations.

We have been introduced to factors in the first `R` tutorial. Factors are a special data structure that is designed to work with categorical data. 

Instead of storing the levels of a categorical variable as text, factors map each level to a unique integer and save the associated labels. This allows for efficient computation since we are using less memory to store data.


## Creating Factors

Let's review how to create a factor vector. We can create a factor by passing a numeric or character vector to the `factor()` function. To assign an ordered label to each level of our data, we include the optional arguments, `levels` and `labels` inside of our `factor()` function.

The example below creates a `weekday_factor` variable from a character vector of day abbreviations. 


```{r}
weekday_factor <- factor(c('M', 'T', 'W', 'Th', 'F', 'M', 'W'),
                        levels = c('M', 'T', 'W', 'Th', 'F'), 
                        labels =  c('Monday', 'Tuesday', 'Wednesday',
                                    'Thursday', 'Friday'))
```



The ordering of the levels in our `weekday_factor` variable is determined from the `levels` order we entered when creating the factor.

`R` functions that work with factors, such as `summary()`, will order their results based on the levels of a factor. For example, when we execute `summary(weekday_factor)` below, the `summary()` function tallies the number of each level that was observed in the factor vector. The results are displayed by the order of the factor levels by default.



```{r}
summary(weekday_factor)
```


## Reversing Levels

What if we would like to reverse the order of the levels within a factor? One way would be to re-create the factor vector with the `factor()` function and the original input. However, this is very time consuming. 

To quickly achieve this task, we can use the `fct_rev()` function from the `forcats` package. This function takes a factor vector as input and automatically reverses the order of the levels.

Let's see how this works with the `weekday_factor` vector. 


```{r}
weekday_reversed <- fct_rev(weekday_factor)

summary(weekday_reversed)
```


```{r}
# Check level ordering
levels(weekday_reversed)
```

Let's see an example of using this function within a `dplyr` analysis workflow. Will will create a new variable, `business_travel_rev` in the `employee_data` data frame that has the reverse ordering of the `business_travel` factor column. The reason we would want to do this may be for how the levels are displayed in a plot or as the results of the `count()` function.


```{r}
# Count factor values of business_travel
employee_data %>% 
  count(business_travel)
```


Next, we reverse the factor levels and the `count()` function displays the results with this new ordering. This will be important for when we learn how to plot data frames with `ggplot` in the coming lessons.


```{r}
employee_data %>% 
  mutate(business_travel_rev = fct_rev(business_travel)) %>% 
  count(business_travel_rev)
```


## Recoding Levels

Another common task is to recode a given factor vector. The `fct_recode()` function is used for this task. This function takes a factor vector as input and then a sequence of assignments in the form `'new_label' = 'old_label'` separated by commas.

Let's say that we wanted to recode our weekday factor and change `Monday` to `Here we go again` and `Friday` to `Party time`. The code below shows how to do this with `fct_recode()`.


```{r}
weekday_recode <- fct_recode(weekday_factor, 
                             'Here we go again' = 'Monday', 
                             'Party time' = 'Friday')

summary(weekday_recode)
```


In the example below, we'll create a `job_level_new` column in the `employee_data` by recoding `job_level`.


```{r}
employee_data <- employee_data %>% 
                 mutate(job_level_new = fct_recode(job_level,
                                                   'Manager' = 'Manager',
                                                   'Manager' = 'Senior Manager'))

# View original level counts
employee_data %>% count(job_level)
```


```{r}
# Recoded factor level counts
employee_data %>% count(job_level_new)
```

## Collapsing Levels

Another type of transformation is collapsing multiple levels of a factor into a smaller subset of levels. This can be done with `fct_collapse()`.
The syntax of `fct_collapse()` is similar to `fct_recode()` except the right side of the `=` can be a character vector of levels to collapse.

The code below demonstrates this function on the `job_level` column from `employee_data`.


```{r}
employee_data %>% 
  mutate(job_level_new = fct_collapse(job_level,
                                      'Below Director' = c('Associate', 'Manager', 'Senior Manager'),
                                      'Director or above' = c('Director', 'Vice President'))) %>% 
  count(job_level_new)
```


## Combining Infrequent Levels

An important transformation for machine learning is combining factor levels with low prevalence in a data set. Using infrequent factor levels to train machine learning algorithms can cause a phenomenon known as over-fitting, where the algorithm is memorizing the patterns in the data.

The `fct_lump_n()` and `fct_lump_prop()` functions are used to combine infrequent levels into a single new category.

`fct_lump_n()` lumps all levels except for the `n` most frequent (or least frequent if n < 0)

`fct_lump_prop()` lumps levels that appear fewer than a specified proportion of times, `prop`

The code below demonstrates both of these functions on the `business_travel` column of `employee_data`.


```{r}
# Original factor levels
employee_data %>% count(job_level)
```


```{r}
# Lump all levels except the top 3
employee_data %>% 
  mutate(job_level_lump_n = fct_lump_n(job_level, n = 3, 
                                             other_level = 'other')) %>% 
  count(job_level_lump_n)        
```



```{r}
# Lump all levels below a proportion of 0.25 (25%)
employee_data %>% 
  mutate(job_level_lump_prop = fct_lump_prop(job_level, prop =  0.25,
                                                   other_level = 'other')) %>% 
  count(job_level_lump_prop)        
```


## Reordering Levels

Many times we would like to reorder the levels of a factor based off the values of another variable in our data set. 

Let's look at an example. The code below shows that the current levels of the `department` variable in `employee_data` are Marketing, Sales, Research, Product Development, IT and Analytics, and Finance and Operations.


```{r}
levels(employee_data$department)
```


What if we need to reorder the levels of `department` based on the median `salary` of each department?

First, let's use `dplyr` to calculate the median `salary` by department and see what the new ordering of levels should be.


```{r}
employee_data %>% group_by(department) %>% 
                  summarise(median_salary = median(salary)) %>% 
                  arrange(median_salary)
```


Next, we could manually recode our `department` variable with this new ordering. However, `forcats` has a function to do this automatically for us.

The `fct_reorder()` takes three important argument:



- `.f` - a factor vector
- `.y` - a numeric vector by which to order the levels of `.f.` (ascending order by default)
- `.fun` - the summary function to apply to `.y` within each level of `.f` (`median` by default)


The code below creates a new variable in `employee_data` which reorders the levels of `department` by the median `salary` values. The `count()` function automatically displays the levels of this new variable by the order we calculated manually in the previous example.


```{r}
employee_data %>% 
  mutate(department_reorderd = fct_reorder(department,
                                           salary)) %>% 
  count(department_reorderd)
```

To order the levels in descending order, just pass `-salary` to the second argument of `fct_reorder()`.


```{r}
employee_data %>% 
  mutate(department_reorderd_desc = fct_reorder(department, -salary)) %>% 
  count(department_reorderd_desc)
```


We can also order the levels by a different summary function. This could be a custom function which we have written or another built-in summary function such as `sd()`.

In the code below, we reorder the levels of `department` by the standard deviation of `salary` within each department.


```{r}
employee_data %>% 
  mutate(department_reorderd_sd = fct_reorder(department, salary, sd)) %>% 
  count(department_reorderd_sd)
```

# Test Your Knowledge

As an exercise, use `dplyr` to calculate the standard deviation of `salary` by `department` in `employee_data`. Does the order in the `department_reordered_sd` variable from above match your results?

```{r}

```


For more on working with factors in `R`, see [R for Data Science - Factors](https://r4ds.had.co.nz/factors.html)


# Dates

In this section we will cover the basics of working with dates in `R`. We will be using the `lubridate` package to create dates, extract components such as the year or month from a date, and basic date arithmetic.


## Creating Dates

Let's start with how to create a date value in `R`. We will be using the `as_date()` function from `lubridate` to achieve this. This function takes a character representation of a date and an optional `format` argument which is a character combination of the symbols below.

`R` recognizes the following date format symbols. See the examples below of how to use these symbols when creating dates.




| **Format Symbol** |               **Meaning**              |      **Example Values**     |
|:-----------------:|:--------------------------------------:|:---------------------------:|
|       %Y          |              4-digit year              |       1996, 2002, 2010      |
|       %y          |              2-digit year              |          96, 02, 10         |
|       %B          |           Unabbreviated month          |   January, April, October   |
|       %b          |            Abbreviated month           |        Jan, Apr, Oct        |
|       %m          | Numeric month (Can include starting 0) |   01 or 1, 03 or 3, 11, 12  |
|       %A          |          Unabbreviated weekday         | Monday, Wednesday, Saturday |
|       %a          |           Abbreviated weekday          |        Mon, Wed, Sat        |
|       %d          |  Numeric day (Can include starting 0)  |   01 or 1, 09 or 9, 15, 30  |



### Dates From Characters

The format symbols from above can handle most use-cases of creating dates from character values. See the examples below.


```{r}
# Most common format
as_date('2020-04-12', format = '%Y-%m-%d')
```


```{r}
# With / instead of -
as_date('2020/04/12', format = '%Y/%m/%d')
```


```{r}
# Another common format
as_date('Apr-12-20', format = '%b-%d-%y')
```


```{r}
# Another common format
as_date('Sunday, Apr 12 2020', format = '%A, %b %d %y')
```


The `as_date()` function will also work on a vector of character values.


```{r}
as_date(c('2020-Feb-15', '2020-Mar-11', '2018-Sep-14'),
        format = '%Y-%b-%d')
```



The `as_date()` function is not able to handle all date formats, however. There is no `R` format symbol for days represented as 31st, for example. When we try to use `as_date()` on this type of input, we get an `NA` value.


```{r}
as_date('Apr 12th 2020', format = '%b-%d-%Y')
```


### Helper Functions

The `lubridate` package has three useful helper function, `ymd()`, `mdy()`, and `dmy()`, for converting character or numeric values into dates. The `y`, `m` and `d` stand for year, month , and day respectively. These functions have built-in formatting checkpoints that intelligently convert most date formats correctly.

See the examples below for how to use these functions for various inputs.



The functions are very resilient to different formatting types, as long as the sequence of year, month, and day match. 

In the code below, we first create a character vector, `chr_dates` with dates in different formats. All of the dates follow the y-m-d pattern, however. In this case, we can use the `ymd()` function.


```{r}
chr_dates <- c('2020-04-12', '2020-Jun-12', 
               '2020-Apr-12th', '18-January-5th',
               '2017/Apr/28')

ymd(chr_dates)
```


Another example using `mdy()`.


```{r}
chr_dates_mdy <- c('April 12th 2017', 'Sep-05-2018', 'dec-23-2019')

mdy(chr_dates_mdy)
```


### Handling Numeric Date Conversions

Sometimes dates are stored as numbers which need to be converted to a date. The helper functions from `lubridate` can also handle these cases.


```{r}
numeric_dates <- c(20200218, 20180412, 20190705)

ymd(numeric_dates)
```



We have a variable in our `home_sales` data frame, `selling_date`, which is a date stored as number. It follows the year-month-day pattern. Let's had a look at the first couple of values. Notice that `selling_date` has `dbl` data type


```{r}
home_sales %>% 
  select(1:3) %>% 
  slice(1:5)
```


Let's convert `selling_date` to a date variable. Notice that the updated `selling_date` column now has a `date` data type with standardized formatting.


```{r}
home_sales <- home_sales %>% 
               mutate(selling_date = ymd(selling_date))

home_sales
```


Common summary functions such as `min`, `median`, `max`, and `range` work with date inputs.


```{r}
home_sales %>% 
  summarise(min_selling_date = min(selling_date),
            median_selling_date = median(selling_date),
            max_selling_date = max(selling_date))
```


## Extracting Date Components

The `year()`, `month()`, and `day()` function can be used to extract components of a date variable.  

These functions take a date variable as input and return the specified component as a number. 

In the example below, we create three new variables in `home_sales`, `selling_month`, `selling_day`, and `selling_year` using these functions. 

The code below also introduces the `relocate()` function from `dplyr` which allows use to reorder columns within a data frame. It has two optional arguments, `.before` or `.after` to determine where to place the specified columns.



```{r}
home_sales <- home_sales %>% 
              mutate(selling_month = month(selling_date),
                     selling_day = day(selling_date),
                     selling_year = year(selling_date)) %>% 
              relocate(selling_month, selling_day, selling_year, 
                       .after = selling_date)

# View results
home_sales
```



## Date Arithmetic 

Another common date operation is to calculate the difference between dates in various units, such as seconds, days, or years. `R` has a built-in function for this task, `difftime()`.

This function takes the following arguments:



- `time1` - a date or datetime value
- `time2` - a date or datetime value
- `tz` - an optional timezone
- `units` - a character of either 'auto', 'secs', 'mins', 'hours', 'days', 'weeks'


Unfortunately, `difftime` doesn't have a 'years` unit, but we can just divide the number of days by 365.25 to obtain this. 


In the example below we have two date vectors, one with the birth dates of students and one with their registration dates for a class. 

Our goal, is to calculate each student's age on their registration date.

```{r}
birth_dates <- ymd(c(19890528, 19960415, 20000815))

registration_dates <- ymd(c(20180914, 20190505, 20200311))

# View dates
birth_dates

registration_dates
```


We can pass our date vectors into the `difftime()` function. We select units of 'days' in this example. We get a numeric vector as output, which we store as `days_diff`. This vector is actually a *difftime* object with a special 'units' attribute. 


```{r}
days_diff <- difftime(registration_dates, birth_dates, units = 'days')

str(days_diff)
```


When doing calculations with dates, I generally recommend converting difftime objects to numeric vectors with `as.numeric()`. This will drop any additional attributes and will make your results compatible with other `R` functions.

```{r}
# Calculate age

age <- as.numeric(days_diff/365.25)

age
```


To calculate a truncated age, you can use the `floor()` function which drops any decimals.


```{r}
floor(age)
```



The example shows how to do these steps with `%>%` operators.


```{r}
(difftime(registration_dates, birth_dates, units = 'days')/365.25) %>% 
  as.numeric() %>% 
  floor()
```

The example below shows how this can be done within a data frame using `dplyr`.


```{r}
my_df <- tibble(birth_date = ymd(c(19890528, 19960415, 20000815)),
                registration_date = ymd(c(20180914, 20190505, 20200311)))

my_df
```



```{r}
my_df %>% 
  mutate(age = (difftime(registration_date, birth_date, units = 'days')/365.25) %>%
                as.numeric() %>% 
                floor())
```


For more examples of working with dates, please see [R for Data Science - Dates](https://r4ds.had.co.nz/dates-and-times.html)



# Text Data

Data that is stored as raw text in character vectors is often referred to as **string** data. This type of data is becoming more prevalent as fields such as [Natural Language Processing](https://en.wikipedia.org/wiki/Natural_language_processing) and [Text Mining](https://en.wikipedia.org/wiki/Text_mining) have become more utilized in machine learning applications.  

The `stringr` package from `tidyverse` contains numerous functions for creating, combining, counting, detecting, and replacing patterns within strings.

In `R` text data is stored in character vectors. These can be created with either single quotes, such as `c('text example')`, or double quotes, `c("another text example")`



## Amazon Shoe Reviews

The code below imports a data frame that contains open-ended customer reviews of various shoes sold on Amazon. The `review` column contains the customer review for a particular product. 

We will be using this data set to showcase the capability of `stringr` to process text data for insights.



```{r}
amazon_reviews <- read_rds(url('https://gmudatamining.com/data/amazon_shoe_reviews.rds'))

amazon_reviews
```


## Combining Strings

All string functions in `stringr` begin with `str_`. In this section we will use the `str_c()` function to combine the elements of different character vectors. 

The `str_c()` function takes the following arguments:



- a sequence of character vectors separated by commas (vectors do not have to be the same length)
- `sep` - the character to separate elements after joining
- `collapse` - the character to separate elements after collapsing


First let's create three character vectors with the first, middle, and last names of individuals.



```{r}
first_name <- c('Lucie', 'Gregg', 'Alex', 'Heather')

middle_initial <- c('A', 'D', 'J', 'K')

last_name <- c('Smith', 'Stevens', 'Gonzalez', 'Chan')
```



### Collapsing Elements

The `str_c()` function can be used to collapse the elements of a character vector into a single element. 

To accomplish this, we pass a character vector into `str_c()` followed by the argument `collapse = your_separating_character`. 



```{r}
# Separate each name with a comma
str_c(first_name, collapse = ', ')

# Separate each name with a space
str_c(first_name, collapse = ' ')
```



### Joining Character Elements

We can also use `str_c()` to combine multiple character vectors of possibly different lengths.

For example, suppose we would like to add 'First Name:' to the beginning of each name in the `first_name` vector. We simply add this string as the first argument to `str_c()`. In the example below, we also add a separating space.

```{r}
str_c('First Name:', first_name, sep = ' ')
```



We can also combine two or more vectors with the same number of elements.



```{r}
str_c(first_name, middle_initial, last_name, sep = ' ')
```


# Test Your Knowledge

How would you adjust the code above to achieve the results below? **Hint**: you can use `str_c()` within the `str_c()` function.


```{r echo = FALSE}
str_c(first_name, str_c(middle_initial, ".", sep = ''), last_name, sep = ' ')
```




## Extracting Subsets

The `str_sub()` can be used to extract a subset of characters from each element of a string. The `str_sub()` function takes a character vector as the first argument followed by the `start` and `end` positions. The `start` and `end` positions can be negative. A `-1` represents the last position, `-2` second to last and so forth.



```{r}
# First character from each first name
str_sub(first_name, 1, 1)
```



```{r}
# Second through fourth characters
str_sub(first_name, 2, 4)
```



```{r}
# Last two characters from each first name
str_sub(first_name, -2, -1)
```



## String Length

The `str_length()` function takes a character vector as an argument and returns the total number of characters within each element. The length calculation includes non-alpha numeric characters such as spaces (' '), commas (','), and other symbols.



```{r}
# Count the number of letters in each first name
str_length(first_name)
```


This function is useful for selecting text by their length. For example, if we wanted to look at short reviews in our Amazon data, we could create a `review_length` column and `filter()` based on its values.

```{r}
amazon_reviews %>% 
  mutate(review_length = str_length(review)) %>% 
  select(review_id, review_length)
```


Let's look at reviews with less than 25 characters.



```{r}
amazon_reviews %>% 
  mutate(review_length = str_length(review)) %>% 
  filter(review_length < 25) %>% 
  select(review_length, review)
```





## Counting Character Sequences

The `str_count()` is useful for counting the occurrences of specific characters or sequences within a string. For example, we could use the `str_count()` function to count the number of times the letter "e" appears in each element of the `first_name` vector.

The `str_count()` function takes a character vector as the first argument followed by a character sequence or regular expression (more on this in the next section).



```{r}
first_name 

# Tally the number of 'e's in each name
str_count(first_name, 'e')
```



Keep in mind that the character match is case-sensitive.



```{r}
# 'A' not included
str_count(first_name, 'a')
```



To demonstrate the usefulness of this function, let's use it to answer the following question on our Amazon reviews data.

How many reviews have at least two occurrences of the word 'great'?

To answer this question, we will implement the following `dplyr` pipeline:

- start with `amazon_reviews`
- create a new column `great_count` which counts the occurrences of 'great' in each `review`
- select the rows where `great_count` is greater than or equal to 2



```{r}
amazon_reviews %>% 
  mutate(great_count = str_count(review, 'great')) %>% 
  filter(great_count >= 2) %>% 
  select(review)
```



## Detecting Patterns

In text analytics, searching for patterns within raw character data is a common task. For example, what if we wanted to find all shoe reviews that contain the word 'awesome' and common variants.

The language of pattern matching in strings is known as **regular expressions** or `regex` for short. In this section, we will go over the basics of pattern matching with `regex`.

For students interested in text analytics, I highly recommended completing the [RegexOne Tutorial](https://regexone.com/). This a comprehensive introduction to regular expressions and can be completed in about an hour.



As a motivating example, we can use regular expressions to find all mentions of `awesome` in the example below.

The `str_view()` function can be used to test matches of regular expressions on a character vector.

This function takes a charcater vector as the first argument and a regular expression as the second. A regular expression is nothing more than a sequence of characters and special symbols encolsed by double quotes (`"`).


```{r}

awesome_vector <- c('awesome shoes', 'Awesome price!', 
                    'awweessooommeee!')

# View regex matches
str_view(awesome_vector, "[Aa]w*e*s*o*m*e*")
```


### Matching Exact Strings

The simplest regex patterns are exact strings, such as words, phrases, or numbers. 

To demonstrate these type of matches, let's create a vector with some reviews.


```{r}
example_reviews <- c('Amazing!', 'Grrrreeaatt fit', 
                     'exelent product...great fit', 
                     'excelent!!!','amazing shoes, small fit',
                     'Perfect 10!', 'Amazing')

example_reviews
```


Below are some examples of matching exact strings. By convention, all regular expression patterns are entered between double quotes. This makes it easier to search for single quotes (`'`) within strings.




```{r}
# Find all instances 'fit'
str_view(example_reviews, "fit")

```




```{r}
# Find all instances '10'
str_view(example_reviews, "10")
```



### Any single character

The `.` symbols is used to represent any character value. For example, `10.` will match the number 10 followed by any single character.


```{r}
# Find all instances of '10' followed by any single character
str_view(example_reviews, "10.")
```



```{r}
# Find all three character sequences with 'x' in the middle
str_view(example_reviews, ".x.")
```



what if we need to match an actual period (`.`) in our text. To do so, we must escape the `.` symbol with a double backslash (`\\`). This tells our regex to match the character `.`


```{r}
# Find all three character sequences with 'x' in the middle
str_view(example_reviews, "\\.")
```


### Optional Characters

The `?` symbol allows us to optionally include the character to the left of the symbol. For example, `great!?` will match both `great` and `great!` since we have indicated that the `!` character is optional in our regex.

As with the `.` symbol, to match a literal `?` in our text, must must escape it with a double backslash (`\\?`)


```{r}
# Find 'Amazing' with optional ! at the end
str_view(example_reviews, "Amazing!?")
```



### Repetition

The `*` and `+` symbols are used to match repetitions of a character.


- `*` will match 0 or more instances of the character to its left
- `+` will match 1 or more instances of the character to its left


For example, the regex `a*b*` will match `aaaa`, `abbb`, `aaabbbbbbb`, `bbbb`. 

However, `a+b+` will only match to `abbb` and `aaabbbbbbb`. 

the difference is that `+` requires the character to be present in the string at least once.



```{r}
# Find multiple periods
str_view(example_reviews, "\\.+")
```



```{r}
# Find 'Great' with possible repeats
str_view(example_reviews, "G+r+e+a+t+")
```



### Anchors

By default, regular expressions will match any part of a string. Many times we need to match patterns at the beginning or end of a string. There are two regex symbols that we can use for this task:

- `^` to match the start of a sequence of characters (placed to the left of the sequence)
- `$` to match the end of of a sequence of characters (placed to the right of the sequence)



```{r}
# Reviews that start with 'A'
str_view(example_reviews, "^A.*")
```



```{r}
# Reviews that end with 'fit'
str_view(example_reviews, "fit$")
```



### Character Classes

If a group of characters are placed inside `[]`, then the regex will match any of those characters. For example, `[Aa]bc` will match either `Abc` or `abc`. 

Any sequence placed inside `[]` is known as a character class. These can be created on the fly, but there are a number of predefined classes that are useful for matching common string patterns.


 - `[\\d]` will match any digit
 - `[a-zA-Z]` will match any letter
 - `[a-zA-Z0-9]` will match any alpha-numeric character

 
Let's find all digits in our example reviews. We will include a `+` after our character class to include multiples.


```{r}
# All non alpha-numeric symbols with repetition
str_view(example_reviews, "[\\d]+")
```


Placing a `^` at the beginning of a character class will match to all characters except for the first instance of the sequence within the class.


```{r}
# Anything that is not a number
str_view(example_reviews, "[^\\d]+")
```



### OR Operator

The `|` symbol represents the logical OR operation within a regex. For example, `amazing|great` will match the string `amazing` or `great`.


```{r}
# Anything that is not a number
str_view(example_reviews, "[Aa]mazing|[Gg]reat")
```



### Filtering a Data Frame with Regular Expressions

The `str_detect()` function from `stringr` is used to detect the presence of regular expressions within a character vector. This function will return a sequence of `TRUE`/`FALSE` values as a results.

For example, the code below checks for the presence of a digit within each element of `example_reviews`.


```{r}
example_reviews

str_detect(example_reviews, "[\\d]")
```


A common data analysis task is to combine `str_detect()` within `filter()` to retrieve rows that match a certain regular expression pattern.

In the example below, we select reviews  from our `amazon_reviews` data with less than 20 characters and at least one instance of the word 'great', with either a capital or lower case 'g' with possible repetition of letters.


```{r}
amazon_reviews %>% 
  filter(str_length(review) < 20, str_detect(review, "[Gr]r+e+a+t+")) %>% 
  select(review)
```


Regular expressions are extremely powerful and becoming an in-demand skill for analytics professionals. A great property of regular expressions is that they are fairly standardized across all programming languages.

For those of you who would like to learn more, head over to the [RegexOne Tutorial](https://regexone.com/) and [R for Data Science - Strings](https://r4ds.had.co.nz/strings.html).


# Iteration

Iteration is the process of applying a function to multiple elements of a collection. A collection could be the elements of a list or the columns of a data frame.

Iteration is so common in data analysis that the `tidyverse` package contains a specialized package, `purrr`, built specifically to make these operations easier. 




## The `map()` Functions

The `map()` function within the `purrr` package can be used to execute FOR loops in an elegant way. The `map()` function always returns a list of results. If the iteration results in a vector, you can control the type of vector output with `map_dbl()`, `map_chr()`, `map_int()`, and `map_lgl()`. However, if the results of the function call do not match the map_*type* call, `purrr` will give an error.



The important arguments to this function are:

- `.x` - an `R` object which we wish to iterate over. Can be a vector, list, or data frame

- `.f` - the function to apply to each element of `.x`. Can be a named function (such as `mean`) or user defined



### Iterating Over a Vector

In the example below, we will apply the `ceiling()` function (which rounds up to the nearest integer) to all elements of `num_vector`


```{r}

num_vector <- c(2.34, 5.45, 6.75)

```


```{r}
# Map the ceiling() function to each element of num_vector
# The map() function always returns a list
map(.x = num_vector, .f = ceiling)

```




```{r}
# Store results in a numeric vector instead of the default list
# .x and .f do not have to be named explicitly
map_dbl(num_vector, ceiling)

```




### Iterating Over a List

The example below applies the `max()` function to each element of `my_list`.


```{r}

my_list <- list(vec_1 = c(1, 4, 6), 
                vec_2 = c(3, 1), 
                vec_3 = c(1, 10, 9, 8))
```



```{r}
max_values <- map_dbl(my_list, max)

```



```{r}
# View results - a named numeric vector
max_values

```




### Iterating Over a Data Frame

The `map()` functions can also be used to iterate over columns of a data frame. 

The example below creates a named character vector that contains the class of each column in the `iris` dataset. The `class` function is applied to each column and returns the data type of the column values.


```{r}
iris_col_types <- map_chr(iris, class)

```



```{r}
# View results
iris_col_types

```



It is also possible to write custom functions within a call to the `map()` functions. Below are two examples of applying a custom function to the numeric columns of the `iris` data frame.



```{r}
# Function that subtracts the median of a vector from all 
# elements and finds the maximum value
max_median_dev <- function(x) {
                 max(x - median(x))
}

```



```{r}
# Apply this function to first 4 columns of iris
# Store output as a numeric vector
map_dbl(iris[ , 1:4], max_median_dev)

```




An equivalent way of achieving this would be with the following syntax within the argument to `.f`. We can use `.x` to represent the variable of iteration and write custom functions directly within the input to `.f`

function(x) { x - mean(x) } is the same as `~ .x - mean(.x)`


```{r}

map_dbl(iris[ , 1:4], ~ max(.x - median(.x)))

```



# Test Your Knowledge

Fill in the code below and use the `map_dbl()` function to create a numeric vector that contains the column averages of all the variables in the built-in data frame, `mtcars`.


```{r}
# View first couple of rows of mtcars
head(mtcars)

```



```{r}
# Use map_dbl to create a vector of average values
mtcars_col_avg <- map_dbl(mtcars, .f = )
```


### Additional Arguments to Functions

It is also possible to pass additional arguments to functions within `map()`. For example, the `rnorm()` function generates random observations from a normal distribution. Its first, second, and third arguments are `n` (number of observations to generate), `mean` (mean of the particular normal distribution), and `sd` (standard deviation of the particular normal distribution.


```{r}
# Generate 4 observations from Normal with mean = 3, sd = 4
rnorm(4, mean = 3, sd = 4)

```


Let's use `map()` to generate observations of normal distributions with mean = 4 and sd = 6 for various values of `n`, which we store in a vector. The additional parameters are passed to the end of the `map()` function.



```{r}
sample_sizes <- c(2, 4, 5)

```


```{r}
# Sample from Normal distribution for n values in sample_sizes
map(sample_sizes, rnorm, mean = 4, sd = 6)
```