---
title: "Inferential Statistics with R"
author: "David Svancer"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = "")
```


This tutorial will demonstrate how to work with theoretical probability distributions in `R`, generate random data from distributions or data sets, and plot probability distribution curves. We will be working with the `diamonds` data frame which is loaded with the `tidyverse` package. Please install the `gridExtra` package before executing the `R` code below. This package will allow us to put multiple ggplot objects onto one panel.


# Data

The `R` code below imports the `tidyverse` and `gridExtra` packages. The `gridExtra` is useful for combining multiple plots created with `ggplot` into a single visualization.

We will be working with the built-in `diamonds` data which is automatically loaded with the `tidyverse` package. A row in this dataset represents a diamond with its associated characteristics and measurements.



```{r}
library(tidyverse)
library(gridExtra)

```


The `diamonds` data set contains the prices and other attributes of almost 54,000 diamonds. To learn more about the data, execute `?diamonds` in your `R` console.


```{r}

diamonds

```


# Random Sampling in R

## Sampling From Data Frames

To demonstrate the process of generating a random sample from a data set, we will use the `diamonds` data frame from `tidyverse`. There are many reasons why we would be interested in taking a random sample of our data, including working with large data files.

For data with millions or billions of rows, it would better to take a random sample of the data, which preserves the statistical relationships present in the data, for data exploration and visualization.

There are two functions in the `dplyr` package designed for this purpose. We can use the `sample_frac()` function from `dplyr` to randomly select a specified proportion of our data or the `sample_n()` function to randomly select a specified number of rows.

These functions are demonstrated in the code below, where we take 2 random samples from the `diamonds` data.

The important arguments to `sample_frac()` are a data frame and the `size` of the sample, which is a proportion between 0 and 1. 

In the code below, we randomly sample 10% (about 5,400) of the rows in `diamonds`


```{r}
# diamonds_1 with sample_frac
diamonds_1 <- sample_frac(diamonds, size = 0.10)

```


The `sample_n()` function has the same first two arguments, however `size` is now an integer between 1 and the total number of rows in the data frame. 

Below we randomly sample 5,400 rows from `diamonds`


```{r}

diamonds_2 <- sample_n(diamonds, size = 5400)

```


Next, let's create 3 plots of the same type from the 3 data sets, `diamonds`, `diamonds_1`, and `diamonds_2`, and arrange them next to each other with the `grid.arrange()` function from the `gridExtra` package.

`ggplot` objects can be saved just like any other `R` objects. In the code below, we save the plots as `original`, `sample_1`, and `sample_2`.

As you can see, the relationship between `price` and `carat` remains approximately the same even with just 10% of the data.

For large datasets, it's best to use a small value of `alpha` within `geom_point()` due to overlapping points.


```{r}

original <- ggplot(data = diamonds, mapping = aes(x = carat, y = price)) + 
            geom_point(alpha = 0.1) + labs(title = "Original", y = 'Diamond Price') + theme_light()

sample_1 <- ggplot(data = diamonds_1, mapping = aes(x = carat, y = price)) + 
            geom_point(alpha = 0.1) + labs(title = "Sample 1", y = '') + theme_light()

sample_2 <- ggplot(data = diamonds_2, mapping = aes(x = carat, y = price)) + 
            geom_point(alpha = 0.1) + labs(title = "Sample 2", y = '') + theme_light()

```


```{r fig.width = 9}
# Arrange the saved plots next to each other
grid.arrange(original, sample_1, sample_2,nrow = 1)

```


## Sampling From Vectors

The `sample()` function takes a sample of the specified size from the elements of a vector either with or without replacement. This is useful for simulating probability experiments, such as tossing a coin or die, and splitting a data set into a training set and test set (more on this when we study cross-validation to access model under-fitting/over-fitting).

Arguments to the `sample()` function:


-   `x` - A (numeric, character or logical) vector of more than one element from which to choose

-   `size` - Non-negative integer giving the number of items to choose

-   `replace` - Should sampling be with replacement? (Default is FALSE)

-   `prob` - A vector of probability weights for obtaining the elements of the vector being sampled. If `NUL`L, elements of `x` are assumed to be equally likely to be chosen



### Tossing a Fair Coin

To demonstrate how `sample()` works, let's simulate tossing a fair coin. The code below produces a single random coin flip where "H" represents lands on heads and "T" represents lands on tails.


```{r}

sample(c("H", "T"), size = 1, prob = c(1/2, 1/2))

```


The code below produces 10 random coin flips. In this case we must use `replace = TRUE` so that after a "heads" is observed it can be sampled again, for example. 

Note that the default of the `prob` argument is to make the elements of the input vector equally likely. Therefore, we do not need to add `prob = c(1/2, 1/2)` since this is automatically done for us. 


```{r}
# A fair coin
sample(c("H", "T"), size = 10, replace = TRUE)

```


To flip a biased coin, we need to specify the unequal probabilities.

```{r}
# A biased coin
sample(c("H", "T"), size = 10, replace = TRUE, prob = c(0.05, 0.95))

```


### Setting Random Seeds

The `set.seed()` function is used to set the random number generator within `R` to a fixed initial starting point. 

As before, we will get random observations from an experiment when we execute any function that performs random sampling.

However, after the first time we execute the function(s), we will get the same results for every successive run. 

This is important if you need to reproduce your results in follow-up analyses. The code below produces a random coin flip that lands on tails. When the code is executed again, it will always land on tails.


```{r}
# Place any integer into set.seed() to initialize
set.seed(314)

sample(c("H", "T"), size = 1)

```


We can also write a function that when called, will produce a random coin flip. In this case, our function will have no formal arguments, it will just execute the code within the body of the function.


```{r}
# Function to generate a random coin flip
coin_flip <- function() {
               sample(c("H", "T"), size = 1) }
```


```{r}
# Call the function 
coin_flip()

```


# Probability Distributions

When working with probability distributions in `R`, we are generally concerned with the following activities:


- Obtaining random samples from a distribution - functions will begin with `r`

- Obtaining quantiles (percentiles) from a distribution - functions will begin with `q`

- Obtaining the probability that a random observation will be less than or equal to a specified value - functions will begin with `p`

- Obtaining the value of the density function at a particular point - functions will begin with `d`


## Random Samples From Distributions

To generate random observations from either a Normal or t distribution, we use `rnorm()` and `rt()`.


The code below will generate a vector of 10 random observations from a [normal distribution](https://en.wikipedia.org/wiki/Normal_distribution) with mean 60 and standard deviation 5.


```{r}
normal_obs <- rnorm(n = 10, mean = 60, sd = 5)

```


```{r}

normal_obs

```


The code below will generate a vector of 10 random observations from a [t distribution](https://en.wikipedia.org/wiki/Student_t-distribution) with 5 degrees of freedom.


```{r}
# Generate a vector of 10 random observations from a t distribution 
# with 5 degrees of freedom
t_obs <- rt(n = 10, df = 5)

```


```{r}
# View results
t_obs

```

## Quantiles (Percentiles)

To obtain the values that correspond to particular percentiles of the normal or t distribution, we use `qnorm()` and `qt()`.

In the code below, we use `qnorm()` to obtain the 95th percentile value for a normal distribution with mean 60 and standard deviation 5. 

This value is equal to 68.22 and means that 95% of values that are sampled at random from this distribution will be less than or equal to 68.22.


```{r}
p_95th_normal <- qnorm(p = 0.95, mean = 60, sd = 5)

```


```{r}
p_95th_normal

```


Next we use `qt()` to obtain the 95th percentile value for a t distribution with 5 degrees of freedom.


```{r}
p_95th_t <- qt(p = 0.95, df = 5)

```


```{r}
# Check results
p_95th_t

```


## Cumulative Probabilities

To obtain the probability a random observation will be less than or equal to a specific value of the normal or t distribution, we use `pnorm()` and `pt()`.

From the output below, the probability that a random sample from a normal distribution with mean 60 and standard deviation 5 will be less than the value 68.22, is 0.95.


```{r}
pnorm(q = 68.22427, mean = 60, sd = 5)

```


For obtaining cumulative probabilities for t distributions, we use the `pt()` function.


```{r}

pt(q = 2.015048, df = 5)

```


## Probability Density Values

To obtain the probability density values at a specific point, we use `dnorm()` and `dt()`. This returns the height of the normal or t distribution function at the given point.

These functions are mainly used for plotting normal and t distributions. They return the density functional values, f(x), at the input value, x, for a particular distribution.

In the code below, we obtain the height of the normal distribution function with mean 60 and standard deviation 5 at the value 60.


```{r}
dnorm(x = 60, mean = 60, sd = 1)

```


The `dt()` function returns the height of a t distribution with a specified degree of freedom.

```{r}
# Density value at x = 1 for a t Distribution
# with 5 degrees of freedom
dt(x = 1, df = 1)

```


# Density Histograms With Normal Density Estimate

Density histograms are used to study the shape of numerical data and determine whether a particular theoretical probability density function is a good fit to model the probability dynamics of that variable in a population. 

Unlike regular histograms, which provide counts of a numeric variable for a series of bins, density histograms provide the proportion of observations falling within a series of bins.

Below we use `geom_histogram()` to make a histogram of the values in the `table` column of the `diamonds` data.


```{r fig.width=9}
# Density histogram of "table"
ggplot(data = diamonds, mapping = aes(x = table, y = ..density..)) +
geom_histogram(fill = "#006EA1", color = "white", bins = 20) + 
labs(title = "Density Histogram of table (Top Diamond Width Relative to Widest Point)",
      x = "Diamond Table",
      y = "Proportion of Observations")

```


We can overlay a theoretical probability density function over a density histogram to determine if the function serves as a good probability model. This is done with `stat_function()` in `ggplot2`.

**Note**: When overlaying functions onto density histograms, make sure to only add `y = ..density..` into the aesthetic mapping of `geom_histogram()`, otherwise it would be passed into `stat_function()` which would generate an error.


```{r fig.width=9}

ggplot(data = diamonds, mapping = aes(x = table)) +
geom_histogram(mapping = aes(y = ..density..), fill = "#006EA1", 
               color = "white", bins = 30) + 
labs(title = "Density Histogram of table (Top Diamond Width Relative to Widest Point)",
      x = "Diamond Table",
      y = "Proportion of Observations") +
stat_function(fun = dnorm, # fun to calculate density value
              args = list(mean = mean(diamonds$table), # args to pass to dnorm
                          sd = sd(diamonds$table)),
              color = "orange",
              size = 1)

```
